Left-Linear Rewriting in Adhesive Categories

### 2024-07-07 concur/final.tex
typos

So far, most of the theoretical results focused on %what are called
\emph{linear rules}: intuitively, a step is required only to consume,
preserve and generate entities.  However, in some situations it is
also necessary to ``merge'' parts of the state.  This happens
in the context of string diagrams mentioned above, but also in
graphical implementations of nominal calculi where, as a result of
name passing, the received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07}, or in the visual modelling of bonding
in biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms
is at the core of e-graphs (equality graphs), used for
rewrite-driven compiler optimisations~\cite{WNW:egg}: rather than
merging nodes corresponding to equal expressions, the idea 
-- conceptually and, as we will see, also mathematically similar
-- is to maintain an
equivalence %\todo{P: more precisely a congruence}
over the nodes.

Technically, to acquire the expressiveness needed for capturing
fusions, i.e.~the merge of some elements in the state, requires to
move from linear to left-linear rewriting rules. While left-linear
rules have been considered in various instances (mainly in categories
of graphs) and some results concerning the corresponding rewriting
theory have been put forward, %to the best of our knowledge 
the phenomena arising when devising a theory of independence and
concurrency for rewriting with left-linear rules in the general
setting of adhesive and quasi-adhesive categories
%(which capture most ``applications'') 
have not been systematically
explored~\cite{Ehrig1976,EHP:BRfTToHLRS,baldan2011adhesivity},
even if an investigation in the setting of quasi-topoi, considering
non-linear rules yet restricting their applicability,
% of rules and focussing as far as DPO in concerned on quasi-adhesivity, 
has been presented in~\cite{behr2021concurrency,BehrHK23}.

@@

Once some sugar has been added, the coffee is ready for drinking. This is represented by rule $\rho_r$, which check for the presence of sugar, represented by the loop $s$, in the coffee, and bring it to the $r$ state, ready for drinking.

>>>>

So far, most of the theoretical results focused on %what are called
\emph{linear rules}: intuitively, a step is required only to consume,
preserve and generate entities.  However, in some situations it is
also necessary to ``merge'' parts of the state.  This happens
in the context of string diagrams mentioned above, but also in
graphical implementations of nominal calculi where, as a result of
name passing, the received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07}, or in the visual modelling of bonding
in biological/chemical processes~\cite{PUY:MBPE}. A similar mechanism
is at the core of e-graphs (equality graphs), used for
rewrite-driven compiler optimisations~\cite{WNW:egg}: rather than
merging nodes corresponding to equal expressions, the idea 
-- conceptually and, as we will see, also mathematically similar
-- is to maintain an
equivalence %\todo{P: more precisely a congruence}
over the nodes.

Technically, to acquire the expressiveness needed for capturing
fusions, i.e.~the merge of some elements in the state, requires to
move from linear to left-linear rewriting rules. While left-linear
rules have been considered in various instances (mainly in categories
of graphs) and some results concerning the corresponding rewriting
theory have been put forward~\cite{Ehrig1976,EHP:BRfTToHLRS,baldan2011adhesivity}, %to the best of our knowledge 
the phenomena arising when devising a theory of independence and
concurrency for rewriting with left-linear rules in the general
setting of adhesive and quasi-adhesive categories
%(which capture most ``applications'') 
have not been systematically
explored,
even if an investigation in the setting of quasi-topoi, considering
non-linear rules yet restricting their applicability,
% of rules and focussing as far as DPO in concerned on quasi-adhesivity, 
has been presented in~\cite{behr2021concurrency,BehrHK23}.

@@

Once some sugar has been added, the coffee is ready for drinking. This is represented by rule $\rho_r$, which checks for the presence of sugar, represented by the loop $s$, in the coffee, and brings it to the $r$ state, ready for drinking.
